Efficient toolkit implementing best practices for principal component analysis of population genetic data

### 2019-10-29 paper-bedpca.tex
save

\abstract{
Principal Component Analysis (PCA) of genetic data is routinely used to infer ancestry and control for population structure in various genetic analyses. However, conducting PCA in large populations is complicated, and there are several potential pitfalls. These pitfalls include (1) capturing Linkage Disequilibrium (LD) structure instead of population structure, (2) projected PCs that suffer from shrinkage bias when projecting PCA from a reference dataset to another independent dataset, and (3) detecting sample outliers. In this work, we explore these pitfalls of using PCA, and present efficient solutions to these pitfalls. Following applications to diverse datasets, we make recommendations for best practices and implement the proposed solutions in an efficient and user-friendly toolkit for performing these analyses in the R packages bigsnpr and bigutilsr.

@@

[SHOULD ALSO TALK ABOUT PCADAPT AND LD]

@@

\noindent[TODO SECTION ON SAMPLE OUTLIER REMOVAL]

@@

Then, we computed again PCA after restricting the individuals included in the computations: we randomly subsampled UKBB data to use only 10,000 British individuals (out of 431,029) and to 5000 Irish individuals (out of 12,755), while keeping all individuals with other self-reported individuals. 
We further removed all individuals for all pairs of related individuals reported by the UKBB. 
This resulted in 48,942 individuals that we used to compute 50 PCs in less than 3 hours using function \texttt{bed\_autoSVD}.

@@

\section*{Funding}

@@

This research has been conducted using the UK Biobank Resource under Application Number 25589.

\section{Declaraction of Interests}

As competing interest, you can write
Michael Blum is an employee at OWKIN.

~\\ \textit{Conflict of Interest:} none declared.

@@

\begin{figure}[!htpb]
\centerline{\includegraphics[width=0.9\textwidth]{proj1000G-UKBB.png}}
\caption{Principal Component (PC) scores 1 to 16 of the 1000 Genomes project and projected individuals from the UK Biobank.
Black points are PC scores of 1000G individuals used for computing PCA.
Blue points are the 488,371 individuals from the UK Biobank, projected using the Online Augmentation, Decomposition, and Procrustes (OADP) transformation. Note that only 20,000 random individuals are represented in this plot.
\label{fig:proj1000G-5}}
\end{figure}

@@

\begin{figure}[!htpb]
\centerline{\includegraphics[width=0.8\textwidth]{PC-1000G-UKBB.pdf}}
\caption{Principal Component (PC) scores 1 to 20 of the 1000 Genomes (1000G) project using variants matched with the UK Biobank. Colors represent populations of the 1000G data.
\label{fig:PC-1000G-UKBB}}
\end{figure}

\begin{figure}[!htpb]
\centerline{\includegraphics[width=0.8\textwidth]{hist-pval-max.pdf}}
\caption{Maximum p-values based on robust Mahalabonis distances of UK Biobank individuals from each of the 26 1000G populations.
\label{fig:ancestry-pval}}
\end{figure}

@@

% latex table generated in R 3.6.0 by xtable 1.8-4 package
% Sat Oct 19 00:18:24 2019
\begin{table}[ht]
\centering
\caption{Ancestry (left) of POPRES individuals and their matching to 1000G populations (top) by our method. See the description of 1000G populations at \url{https://www.internationalgenome.org/category/population/}.} 
\label{tab:ancestry-pred-popres}
\begin{tabular}{|l|c|c|c|c|c|}
  \hline
 & EUR\_CEU & EUR\_GBR & EUR\_IBS & EUR\_TSI & NA \\ 
  \hline
Anglo-Irish Isles & 159 & 57 &  &  & 50 \\ 
  Belgium & 28 & 3 &  &  & 12 \\ 
  Central Europe & 3 &  &  & 1 & 51 \\ 
  Eastern Europe & 2 &  &  &  & 28 \\ 
  France & 14 &  & 13 &  & 62 \\ 
  Germany & 37 &  &  &  & 34 \\ 
  Italy &  &  & 12 & 19 & 188 \\ 
  Netherlands & 10 & 2 &  &  & 5 \\ 
  Scandinavia & 5 &  &  &  & 10 \\ 
  SE Europe &  &  &  & 5 & 89 \\ 
  SW Europe &  &  & 193 &  & 71 \\ 
  Switzerland & 33 & 1 & 6 &  & 182 \\ 
   \hline
\end{tabular}
\end{table}

% latex table generated in R 3.6.0 by xtable 1.8-4 package
% Sat Oct 19 00:28:24 2019
\begin{table}[ht]
\centering
\caption{Ancestry (left) of Celiac individuals and their matching to 1000G populations (top) by our method. See the description of 1000G populations at \url{https://www.internationalgenome.org/category/population/}.} 
\label{tab:ancestry-pred-celiac}
\begin{tabular}{|l|c|c|c|c|c|c|c|}
  \hline
 & AMR\_MXL & EUR\_CEU & EUR\_FIN & EUR\_GBR & EUR\_IBS & EUR\_TSI & NA \\ 
  \hline
Finland &  & 4 & 1543 & 1 &  & 1 & 922 \\ 
  Italy &  &  &  &  & 15 & 229 & 795 \\ 
  Netherlands &  & 802 &  & 382 & 1 &  & 463 \\ 
  UK & 1 & 2865 & 1 & 2532 & 4 & 5 & 1346 \\ 
   \hline
\end{tabular}
\end{table}

>>>>

\abstract{

[REDO LATER]

Principal Component Analysis (PCA) of genetic data is routinely used to infer ancestry and control for population structure in various genetic analyses. However, conducting PCA in large populations is complicated, and there are several potential pitfalls. These pitfalls include (1) capturing Linkage Disequilibrium (LD) structure instead of population structure, (2) projected PCs that suffer from shrinkage bias when projecting PCA from a reference dataset to another independent dataset, and (3) detecting sample outliers. In this work, we explore these pitfalls of using PCA, and present efficient solutions to these pitfalls. Following applications to diverse datasets, we make recommendations for best practices and implement the proposed solutions in an efficient and user-friendly toolkit for performing these analyses in the R packages bigsnpr and bigutilsr.

@@

[SHOULD ALSO TALK ABOUT PCADAPT -> CONFOUNDED BY LD + SHOULD COMPUTE MAHA ON UNLINKED]

@@

\noindent[TODO SECTION ON SAMPLE OUTLIER REMOVAL / HOMOGENEOUS SAMPLES]

@@

Then, we computed again PCA after restricting the individuals included in the computations: we randomly subsampled UKBB data to use only 10,000 British individuals (out of 431,029) and 5000 Irish individuals (out of 12,755), while keeping all individuals with other self-reported individuals. 
We further removed all pairs of related individuals reported by the UKBB (both individuals in each pair). 
This resulted in 48,942 individuals that we used to compute 50 PCs, which took less than 3 hours using function \texttt{bed\_autoSVD} (that converged after 4 iterations of automatic LD removal). 
We show that we are able to capture at least 40 PCs that display visual population structure (Figures \ref{fig:UKBB-scores2} and \ref{fig:UKBB-screeplot2}).
We then projected all 439,429 remaining individuals from UKBB onto this PCA space in 41 minutes. Note that these individuals are not supposed to be related to any of the 48,942 individuals used for training PCA because we removed both individuals from each pair of related individuals in UKBB. Projection of new individuals show again clear shrinkage when using simple projection (between 1.00 for PC1 and 1.80 for PC50), but no visible bias when using OADP projection (Figure \ref{fig:projUKBB-related}).

@@

@@

Authors thank Rounak Dey and Seunggeun Lee for helpful discussions about PCA projection.
This research has been conducted using the UK Biobank Resource under Application Number 25589.

F.P., J.J.M.\ and B.J.V.\ acknowledge Niels Bohr Professorship from the Danish National Research Foundation to Prof.\ John J.\ McGrath, and the Lundbeck Foundation Initiative for Integrative Psychiatric Research, iPSYCH (R248-2017-2003).

\section*{Declaraction of Interests}

Michael Blum is an employee of OWKIN France.
The other authors declare no competing interests.

@@

\begin{figure}[!htpb]
\centerline{\includegraphics[width=0.9\textwidth]{proj1000G-UKBB.png}}
\caption{Principal Component (PC) scores 1 to 16 of the 1000 Genomes project and projected individuals from the UK Biobank.
Black points are PC scores of 1000G individuals used for computing PCA.
Blue points are the 488,371 individuals from the UK Biobank, projected using the Online Augmentation, Decomposition, and Procrustes (OADP) transformation. 
Note that only 20,000 random projected individuals are represented in this plot.
\label{fig:proj1000G-5}}
\end{figure}

@@

@@

\begin{figure}[!htpb]
\centerline{\includegraphics[width=0.8\textwidth]{UKBB-screeplot-restricted.pdf}}
\caption{Scree plot: plot of singular values computed from the UK Biobank using 48,942 individuals of diverse ancestries.
\label{fig:UKBB-screeplot2}}
\end{figure}

\begin{figure}[!htpb]
\centerline{\includegraphics[width=0.9\textwidth]{UKBB-scores-restricted.png}}
\caption{Principal Component (PC) scores 27 to 50 computed from the UK Biobank using 48,942 individuals of diverse ancestries. Different colors represent different self-reported ancestries.
\label{fig:UKBB-scores2}}
\end{figure}

\begin{figure}[!htpb]
\centerline{\includegraphics[width=0.9\textwidth]{projUKBB-related.png}}
\caption{Principal Component (PC) scores 27 to 50 of the UK Biobank.
Black points are the 48,942 individuals of diverse ancestries used for computing PCA.
Red points are the remaining UKBB individuals, projected by simply multiplying their genotypes by the corresponding PC loadings.
Blue points are the remaining UKBB individuals, projected using the Online Augmentation, Decomposition, and Procrustes (OADP) transformation.
Note that only 20,000 random projected individuals are represented in this plot.
\label{fig:projUKBB-related}}
\end{figure}
